---
title: "hw4"
author: "Andre Gala-Garza"
date: "4/12/2026"
format: html
editor: visual
embed-resources: true
---

**Disclaimer:** Generative AI was used to assist with templating and writing code in this assignment; however, this code was checked manually and edited by hand to ensure accuracy.

**Source:** OpenAI. (2026). *ChatGPT* \[Large language model\]. <https://chatgpt.com/>.

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

options(
  max.print = 100,
  tibble.print_max = 10,
  tibble.width = Inf
)
```

```{r}
library(tidyverse)
library(dplyr)
```

# HPC

## Question 1: fun2

```{r}
# running cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}

fun2alt <- function(mat) {
  k <- ncol(mat)
  ans <- mat
  for (j in 2:k) {
    ans[, j] <- ans[, j] + ans[, j - 1]
  }
  ans
}
```

```{r}
# Use the functions with this data
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)
```

```{r}
# Check that both functions give the same result
out1 <- fun2(dat)
out2 <- fun2alt(dat)

all.equal(out1, out2)
```
```{r}
# Benchmark test
library(microbenchmark)

mb <- microbenchmark(
  original = fun2(dat),
  faster = fun2alt(dat),
  times = 100
)

mb
```
```{r}
library(ggplot2)

mb_df <- as.data.frame(mb)

ggplot(mb_df, aes(x = expr, y = time / 1e6)) +
  geom_boxplot() +
  labs(
    title = "Runtime comparison: fun2 vs fun2alt",
    x = "Function",
    y = "Runtime (ms)"
  )
```
fun2alt() is faster because it keeps only one loop and performs each cumulative update on a whole column instead of one matrix element at a time, while still returning exactly the same output as fun2().

## Question 2: fun3

```{r}
fun3 <- function(x) {
  out <- numeric(length(x))
  cumsum_x <- 0
  for (i in seq_along(x)) {
    if (x[i] == 0) {
      cumsum_x <- 0
    } else {
      cumsum_x <- cumsum_x + x[i] 
    }
    out[i] <- cumsum_x
  }
  out
}

fun3alt <- function(x) {
  cs <- cumsum(x)
  last_zero <- cummax(seq_along(x) * (x == 0))
  offset <- numeric(length(x))
  nz <- last_zero > 0
  offset[nz] <- cs[last_zero[nz]]
  cs - offset
}
```

```{r}
# Small example for easy checking
x_small <- c(3, 2, 1, 0, 5, 4, 0, 2, 2, 1, 0, 7)

# Larger random example
set.seed(2315)
x_big <- sample(c(0, -2:-1, 1:5), size = 100000, replace = TRUE)
```

```{r}
# Check accuracy
fun3(x_small)
fun3alt(x_small)

all.equal(fun3(x_small), fun3alt(x_small))
all.equal(fun3(x_big), fun3alt(x_big))
```

```{r}
# Benchmark test
library(microbenchmark)

mb <- microbenchmark(
  original = fun3(x_big),
  faster = fun3alt(x_big),
  times = 100
)

mb
```

```{r}
library(ggplot2)

mb_df <- as.data.frame(mb)

ggplot(mb_df, aes(x = expr, y = time / 1e6)) +
  geom_boxplot() +
  labs(
    title = "Runtime comparison: fun3 vs fun3alt",
    x = "Function",
    y = "Runtime (ms)"
  )
```
fun3alt() is faster because it replaces the element-by-element R loop with vectorized base R operations (cumsum() and cummax()), which do the work in optimized internal code with less interpreter overhead.

## Question 3: sim_pi

```{r}
library(parallel)

sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Here is an example of the run
set.seed(156)
sim_pi(1000) # 3.132
```

```{r}
# Serial version
# This runs the simulation a 4,000 times, each with 10,000 points
set.seed(1231)

serial_time <- system.time({
  ans_serial <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans_serial))
})
```

```{r}
# Parallel version
n_cores <- max(1, detectCores() - 1)
cl <- makeCluster(n_cores)

clusterSetRNGStream(cl, 1231)
clusterExport(cl, "sim_pi")

parallel_time <- system.time({
  ans_parallel <- unlist(parLapply(cl, 1:4000, sim_pi, n = 10000))
  print(mean(ans_parallel))
})

stopCluster(cl)
```
```{r}
# Visualize that both methods give similar estimates of pi
hist(
  ans_serial,
  breaks = 30,
  freq = FALSE,
  main = "Pi estimates: serial vs parallel",
  xlab = "Estimated pi",
  col = rgb(0, 0, 1, 0.4),
  xlim = range(c(ans_serial, ans_parallel))
)

hist(
  ans_parallel,
  breaks = 30,
  freq = FALSE,
  add = TRUE,
  col = rgb(1, 0, 0, 0.4)
)

legend(
  "topright",
  legend = c("Serial", "Parallel"),
  fill = c(rgb(0, 0, 1, 0.4), rgb(1, 0, 0, 0.4))
)
```
```{r}
# Barplot comparison
runtime_df <- c(
  serial = serial_time["elapsed"],
  parallel = parallel_time["elapsed"]
)

barplot(
  runtime_df,
  main = "Elapsed runtime: serial vs parallel",
  ylab = "Seconds"
)
```
The parallel version is faster because it distributes the 4000 independent simulation runs across multiple CPU cores so they can be computed simultaneously instead of sequentially.

# SQL

```{r}
# install.packages(c("RSQLite", "DBI", "readr"))

library(RSQLite)
library(DBI)
library(readr)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Read in tables
surveys <- read_csv("surveys.csv")
species <- read_csv("species.csv")
plots <- read_csv("plots.csv")

# Copy data.frames to database
dbWriteTable(con, "surveys", surveys, overwrite = TRUE)
dbWriteTable(con, "species", species, overwrite = TRUE)
dbWriteTable(con, "plots", plots, overwrite = TRUE)
```
```{r}
dbListTables(con)
```

## Question 4: Surveys Summary

```{r}
dbGetQuery(con,
  "PRAGMA table_info(surveys)"
)
```
```{r}
dbGetQuery(con,
  "SELECT
    year,
    month,
    day,
    species_id,
    ROUND(weight / 1000.0, 2) AS weight_kg
  FROM surveys
  WHERE year = 1999
  ORDER BY species_id"
)
```


## Question 5: Animal Weights

```{r}
# Weights for all animals in 'surveys'
dbGetQuery(con,
  "SELECT
    year,
    month,
    day,
    species_id,
    ROUND(weight / 1000.0, 2) AS weight_kg
  FROM surveys
  WHERE year = 1999
  ORDER BY species_id"
)
```

## Question 6: Counts and Weights

```{r}
species_year_summary <- dbGetQuery(con, "
  SELECT
    year,
    species_id,
    COUNT(*) AS n_counted,
    AVG(weight) AS avg_weight
  FROM surveys
  GROUP BY year, species_id
  ORDER BY year, species_id
")

species_year_summary
```

## Question 7: Number of Species per Taxa

```{r}
taxa_counts <- dbGetQuery(con, "
  SELECT
    taxa,
    COUNT(*) AS n_species
  FROM species
  GROUP BY taxa
  HAVING COUNT(*) > 10
  ORDER BY n_species DESC
")

taxa_counts
```

## Question 8: Number of Animals per Genus

```{r}
genus_plot_counts <- dbGetQuery(con, "
  SELECT
    s.plot_id,
    sp.genus,
    COUNT(*) AS n_caught
  FROM surveys AS s
  INNER JOIN species AS sp
    ON s.species_id = sp.species_id
  GROUP BY s.plot_id, sp.genus
  ORDER BY s.plot_id ASC, n_caught DESC
")

genus_plot_counts
```

## Question 9: Number of Specimens per Plot

```{r}
species_plottype_counts <- dbGetQuery(con, "
  SELECT
    p.plot_type,
    s.species_id,
    sp.genus,
    sp.species,
    COUNT(*) AS n_captured
  FROM surveys AS s
  INNER JOIN species AS sp
    ON s.species_id = sp.species_id
  INNER JOIN plots AS p
    ON s.plot_id = p.plot_id
  WHERE s.species_id IS NOT NULL
    AND LOWER(sp.species) <> 'unknown'
  GROUP BY p.plot_type, s.species_id, sp.genus, sp.species
  ORDER BY p.plot_type, n_captured DESC, s.species_id
")

species_plottype_counts
```

## Question 10: Average Weight of Rodent Species

```{r}
rodent_avg_weight <- dbGetQuery(con, "
  SELECT
    s.species_id,
    sp.genus,
    sp.species,
    AVG(s.weight) AS avg_weight
  FROM surveys AS s
  INNER JOIN species AS sp
    ON s.species_id = sp.species_id
  WHERE sp.taxa = 'Rodent'
  GROUP BY s.species_id, sp.genus, sp.species
  ORDER BY avg_weight DESC
")

rodent_avg_weight
```

## Question 11: Number of Animals per Genus (ggplotly)

```{r}
library(dplyr)
library(ggplot2)
library(plotly)

genus_plot_counts_1_5 <- genus_plot_counts %>%
  filter(plot_id %in% 1:5)

p <- ggplot(genus_plot_counts_1_5, aes(x = factor(plot_id), y = n_caught, fill = genus)) +
  geom_col(position = "dodge") +
  labs(
    title = "Number of animals caught by genus for plots 1-5",
    x = "Plot ID",
    y = "Number caught",
    fill = "Genus"
  )

plotly::ggplotly(p)
```

## Question 12: Counts and Weights (plotly)

```{r}
library(dplyr)
library(plotly)

species_year_plot <- species_year_summary %>%
  filter(!is.na(avg_weight))

plot_ly(
  data = species_year_plot,
  x = ~year,
  y = ~avg_weight,
  color = ~species_id,
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "Year: %{x}",
    "<br>Species: %{color}",
    "<br>Average weight: %{y:.2f}",
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Average species weight over time",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average weight")
  )
```
One potential solution to remove noise in this plot is to keep only species that appear in at least 10 years:
```{r}
species_year_plot_clean <- species_year_summary %>%
  filter(!is.na(avg_weight)) %>%
  group_by(species_id) %>%
  filter(n() >= 10) %>%
  ungroup()

plot_ly(
  data = species_year_plot_clean,
  x = ~year,
  y = ~avg_weight,
  color = ~species_id,
  type = "scatter",
  mode = "lines+markers",
  hovertemplate = paste(
    "Year: %{x}",
    "<br>Species: %{color}",
    "<br>Average weight: %{y:.2f}",
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Average species weight over time (species with at least 10 years of data)",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average weight")
  )
```

We can also remove the markers and keep only the lines in the plot:
```{r}
plot_ly(
  data = species_year_plot_clean,
  x = ~year,
  y = ~avg_weight,
  color = ~species_id,
  type = "scatter",
  mode = "lines",
  hovertemplate = paste(
    "Year: %{x}",
    "<br>Species: %{color}",
    "<br>Average weight: %{y:.2f}",
    "<extra></extra>"
  )
) %>%
  layout(
    title = "Average species weight over time",
    xaxis = list(title = "Year"),
    yaxis = list(title = "Average weight")
  )
```