---
title: "lab-09"
author: "Andre Gala-Garza"
date: "3/19/2026"
format: html
editor: visual
embed-resources: true
---

**Disclaimer:** Generative AI was used to assist with templating and writing code in this assignment; however, this code was checked manually and edited by hand to ensure accuracy.

**Source:** OpenAI. (2026). *ChatGPT (GPT-5.4 Thinking)* \[Large language model\]. <https://chatgpt.com/>.

```{r}
# install.packages(c("RSQLite", "DBI"))

library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
doctor <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/doctor.csv")
visit <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/visit.csv")
patient <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/patient.csv")
payment <- read.csv("https://raw.githubusercontent.com/dmcable/BIOSTAT620W26/main/data/hospital/payment.csv")

# Copy data.frames to database
dbWriteTable(con, "doctor", doctor)
dbWriteTable(con, "visit", visit)
dbWriteTable(con, "patient", patient)
dbWriteTable(con, "payment", payment)
```

```{r}
dbListTables(con)
```
**Note:** For each query, the LIMIT command was used to prevent the entire output from displaying in the knitted HTML file.

# Exercise 1

```{r}
dbGetQuery(con,
  "SELECT doctor_id, first_name, last_name
  FROM doctor
  ORDER by last_name, first_name
  LIMIT 10"
)
```

# Exercise 2

```{r}
dbGetQuery(con,
  "SELECT doctor_id, first_name, last_name
  FROM doctor
  WHERE last_name IN ('WILLIAMS', 'DAVIS')"
)
```

# Exercise 3

```{r}
dbGetQuery(con,
  "SELECT DISTINCT visit_id
  FROM visit
  WHERE date(visit_start_date) = '2005-07-05'
  LIMIT 10"
)
```

# Exercise 4

## Exercise 4.1

```{r}
dbGetQuery(con,
  "SELECT *
  FROM payment
  WHERE amount IN (199, 799, 999)
  LIMIT 10"
)
```

## Exercise 4.2

```{r}
dbGetQuery(con,
  "SELECT *
  FROM payment
  WHERE amount > 500
  LIMIT 10"
)
```

## Exercise 4.3

```{r}
dbGetQuery(con,
  "SELECT *
  FROM payment
  WHERE amount > 500 AND amount < 800
  LIMIT 10"
)
```

# Exercise 5
```{r}
dbGetQuery(con,
  "SELECT payment_id, amount
  FROM payment
  INNER JOIN patient
  on payment.patient_id = patient.patient_id
  WHERE last_name = 'DAVIS'"
)
```

# Exercise 6

## Exercise 6.1
```{r}
dbGetQuery(con,
  "SELECT COUNT(*)
  FROM visit"
)
```

## Exercise 6.2
```{r}
dbGetQuery(con,
  "SELECT patient_id, COUNT(*)
  FROM visit
  GROUP BY patient_id
  LIMIT 10"
)
```

## Exercise 6.3
```{r}
dbGetQuery(con,
  "SELECT patient_id, COUNT(*)
  FROM visit
  GROUP BY patient_id
  ORDER BY patient_id DESC
  LIMIT 10"
)
```

## Exercise 6.4
```{r}
dbGetQuery(con,
  "SELECT patient_id, COUNT(*)
  FROM visit
  GROUP BY patient_id
  HAVING COUNT(*) >= 40
  ORDER BY patient_id DESC"
)
```

# Exercise 7

```{r}
dbGetQuery(con,
  "SELECT MAX(amount), MIN(amount), AVG(amount), SUM(amount)
  FROM payment"
)
```

## Exercise 7.1
```{r}
dbGetQuery(con,
  "SELECT MAX(patient_id), MIN(patient_id), AVG(patient_id), SUM(patient_id)
  FROM payment"
)
```

## Exercise 7.2
```{r}
dbGetQuery(con,
  "SELECT MAX(patient_id), MIN(patient_id), AVG(patient_id), SUM(patient_id)
  FROM payment
  WHERE patient_id IN (
      SELECT patient_id
      FROM payment
      GROUP BY patient_id
      HAVING COUNT(*) > 5
  )"
)
```

# Exercise 8

## Exercise 8.1
```{r}
dbGetQuery(con,
  "SELECT doctor_id, first_name, last_name, AVG(visit_length)
  FROM (
    SELECT
      visit.doctor_id,
      first_name,
      last_name,
      julianday(visit_end_date) - julianday(visit_start_date) AS visit_length
    FROM doctor
    INNER JOIN visit
    ON doctor.doctor_id = visit.doctor_id
  )
  GROUP BY doctor_id
  LIMIT 10"
)
```

## Exercise 8.2

```{r}
dbGetQuery(con,
  "SELECT doctor_id, first_name, last_name, AVG(visit_length)
  FROM (
    SELECT
      visit.doctor_id,
      first_name,
      last_name,
      julianday(visit_end_date) - julianday(visit_start_date) AS visit_length
    FROM doctor
    INNER JOIN visit
    ON doctor.doctor_id = visit.doctor_id
  )
  GROUP BY doctor_id
  ORDER BY AVG(visit_length) DESC
  LIMIT 5"
)
```

```{r}
dbGetQuery(con,
  "SELECT doctor_id, first_name, last_name, AVG(visit_length)
  FROM (
    SELECT
      visit.doctor_id,
      first_name,
      last_name,
      julianday(visit_end_date) - julianday(visit_start_date) AS visit_length
    FROM doctor
    INNER JOIN visit
    ON doctor.doctor_id = visit.doctor_id
  )
  GROUP BY doctor_id
  ORDER BY AVG(visit_length) ASC
  LIMIT 5"
)
```

```{r}
# clean up
dbDisconnect(con)
```